---
title: "Create Food Security Supplement Cohort"
author: "Dr. Roch Nianogo, Bowen Zhang, Dr. Hua Zhou"
format:
  html:
    theme: cosmo
    embed-resources: true
    number-sections: true
    toc: true
    toc-depth: 4
    toc-location: left
    code-fold: false
    link-external-icon: true
    link-external-newwindow: true
---

```{r}
#| code-fold: true
#| output: false

library(tidyverse)
library(gtsummary)
```

In this tutorial, we ingest the 2021 Food Security Supplement (FSS) data from the Current Population Survey (CPS) and create a cohort of socio-economial variables. We then summarize the characteristics of the cohort and examine the relationship between food security status and household characteristics.

# Ingest and tidy data

We include the following independent variables in the model:

-   HRNUMHOU: Total number of persons living in the household.
-   HRHTYPE: Household type
    -   0 = Non-interview household
    -   1 = Husband/wife primary family (neither AF)
    -   2 = Husband/wife primary family (either/both AF)
    -   3 = Unmarried civilian male-primary. family householder
    -   4 = Unmarried civilian female-prim family householder
    -   5 = Primary family householder-rp in AF, unmarried.
    -   6 = Civilian male primary individual
    -   7 = Civilian female primary individual
    -   8 = Primary individual householder-rp in AF
    -   9 = Group quarters with family
    -   10 = Group quarters without family
-   GEREG: Region
    -   1 = Northeast
    -   2 = Midwest
    -   3 = South
    -   4 = West
-   PRTAGE: Age of the reference person
-   PEHSPNON: Hispanic or non-hispanic of the reference person
    -   1 = Hispanic
    -   2 = Non-Hispanic
-   PTDTRACE: Race of the reference person
    -   1 = White only
    -   2 = Black only
    -   3 = American Indian, Alaskan Native only
    -   4 = Asian only
    -   5 = Hawaiian/Pacific Islander only
    -   6\~26 = Others
-   PEEDUCA: Highest level of education completed
    -   31 = Less than 1st grade
    -   32 = 1st, 2nd, 3rd, or 4th grade
    -   33 = 5th or 6th grade
    -   34 = 7th or 8th grade
    -   35 = 9th grade
    -   36 = 10th grade
    -   37 = 11th grade
    -   38 = 12th grade no diploma
    -   39 = High school grad-diploma or equiv (GED)
    -   40 = Some college but no degree
    -   41 = Associate degree-occupational/vocational
    -   42 = Associate degree-academic program
    -   43 = Bachelor's degree (ex: BA, AB, BS)
    -   44 = Master's degree (ex: MA, MS, MEng, MEd, MSW)
    -   45 = Professional school degree (ex: MD, DDS, DVM)
    -   46 = Doctorate degree (ex: PhD, EdD)
-   PEMLR: Employment status
    -   1 = Employed-at work
    -   2 = Employed-absent
    -   3 = Unemployed-on layoff
    -   4 = Unemployed-looking
    -   5 = Not in labor force-retired
    -   6 = Not in labor force-disabled
    -   7 = Not in labor force-other
-   PRCHLD: Presence of own children \< 18 years old
    -   0 = No children
    -   1\~15 = Have children
-   HRPOOR: Poverty status
    -   1 = Poor (Below 185% poverty)
    -   2 = Not poor (Above 185% poverty or income not reported)

First import data from the Census API:

```{r}
fss_21_data <- censusapi::getCensus(name = "cps/foodsec/dec",
                                    vintage = 2021,
                                    vars = c("HRHHID", "HRHHID2", "HRNUMHOU",
                                             "HRHTYPE", "GEREG", "HRPOOR",
                                             "PRTAGE", "PEEDUCA", "PRCHLD",
                                             "PEMLR", "PTDTRACE", "PEHSPNON",
                                             "HRFS12M1", "HHSUPWGT",
                                             "PERRP")) |>
  as_tibble() |>
  print()
```

Recall that we can look at the details of a particular categorical variable by

```{r}
censusapi::listCensusMetadata(
  name = "cps/foodsec/dec",
  vintage = 2021,
  type = "values",
  variable = "PRCHLD"
  ) |>
  as_tibble() |>
  arrange(as.numeric(as.character(code))) |>
  print()
```

Data tidying:

```{r}
data_clean <- fss_21_data |>
  # Remove cases
  filter(
    PERRP %in% c(40, 41), # Reference person
    HRFS12M1 %in% c(1, 2, 3), # Valid food security status codes
    PEMLR %in% c(1, 2, 3, 4, 5, 6, 7), # Valid employment status codes
    ) |>
  # Data preprocessing
  mutate(
    # Food security status
    HRFS12M1 = factor(HRFS12M1,
                      levels = c("1", "2", "3"),
                      labels = c("Food Security",
                                 "Low Food Security",
                                 "Very Low Food Security")),
    # Number of persons living in the household
    HRNUMHOU = as.numeric(HRNUMHOU),
    # Household weight
    HHSUPWGT = as.numeric(HHSUPWGT),
    # Combining some categories, cut uses (a, b] interval notation
    # Household type
    HRHTYPE = cut(as.numeric(HRHTYPE), breaks = c(0, 2, 5, 8, 10),
                  labels = c("MarriedFamily", "UnmarriedFamily",
                             "Individual", "GroupQuarters")),
    # Region
    GEREG = factor(GEREG, levels = c(1, 2, 3, 4),
                   labels = c("Northeast", "Midwest", "South", "West")),
    # Education level
    PEEDUCA = cut(as.numeric(PEEDUCA), breaks = c(30, 38, 42, 46),
                  labels = c("LessThanHighSchool",
                             "HighSchoolOrAssociateDegree",
                             "CollegeOrHigher")),
    # Employment status
    PEMLR = cut(as.numeric(PEMLR), breaks = c(0, 2, 4, 7),
                labels = c("Employed", "NotEmployed",
                           "NotInLaborForce")),
    # Age of reference person
    PRTAGE = as.numeric(PRTAGE),
    # Hispanic or Non-Hispanic
    PEHSPNON = factor(PEHSPNON, levels = c(2, 1),
                      labels = c("NonHispanic", "Hispanic")),
    # Race of reference person
    PTDTRACE = fct_recode(PTDTRACE,
                          "White" = "1" ,
                          "Black" = "2",
                          "AIAN" = "3",
                          "Asian" = "4",
                          "HPI" = "5") |>
      fct_other(keep = c("White", "Black", "AIAN", "Asian", "HPI")),
    # Presence of children
    PRCHLD = ifelse(PRCHLD == 0, "NoChildren", "Children") |> as_factor() |> relevel(ref = "NoChildren"),
    # Poverty status
    HRPOOR = factor(HRPOOR, levels = c(2, 1),
                    labels = c("NotPoor", "Poor"))
  ) |>
  # variables we want keep in analysis
  select(HRFS12M1, HRNUMHOU, HRHTYPE, GEREG, PRCHLD, HRPOOR,
         PRTAGE, PEEDUCA, PEMLR, PTDTRACE, HHSUPWGT, PEHSPNON) |>
  # binary outcome:
  # `HRFS12M1 = 0` if the household is in food security and `HRFS12M1 = 1` o/w
  mutate(HRFS12M1_binary = fct_recode(HRFS12M1,
                                      "Food Security" = "Food Security",
                                      "Food Insecure" = "Low Food Security",
                                      "Food Insecure" = "Very Low Food Security")) |>
  print()
```

# Numerical summary

Numerical summary:
```{r}
data_clean |>
  tbl_summary(by = HRFS12M1_binary)
```

# Graphical summary


# Save data set

Save as RDS file:
```{r}
write_rds(data_clean, "fss21.rds")
```